---
title: "iClusterVB: A variational Bayes approach - User Manual"
output: github_document
---



__iClusterVB: A Next-Generation R Package for Bayesian Clustering__


Traditional clustering methods grapple with the challenges posed by high dimensionality, multimodality, and computational intricacies. In response to these hurdles, we present iClusterVB—an integrative clustering approach grounded in variational Bayesian inference. iClusterVB allows researchers to navigate high-dimensional datasets comprising mixed-type data, including continuous, categorical, and count variables, as well as multi-view data. With iClusterVB, researchers can harness the power of Bayesian inference to unlock the latent structure within their datasets, paving the way for transformative discoveries in biomedicine and beyond.


## Installing and loading the Package


```{r eval = F}
devtools::install_github("AbdalkarimA/iClusterVB")
```


```{r}
library(iClusterVB)
```
## Simulated Data


The simulated dataset is a list of four seperate datasets, as we might expect from multi-view data. Two of those are continuous (continuous1_data and continuous2_data, such as gene expression data), one is count (count_data, such as sequencing data), and one is binary (binary_data, such as the presence or absence of mutations). Each of the four datasets contains 100 samples and 50 variables. However, only 10 were truly informative and the others were noise. This can be seen in the heatmaps below.


```{r}
load("sim_data.Rdata")
```


### Viewing simulated data: Heatmap for features by clusters

```{r}
library(pheatmap)
mat_colors <- list(Clusters = c("green", "blue", "purple", "red"))
mat_col <- data.frame(Clusters = as.numeric(simdata$cluster))

dat1 <- data.frame(t(simdata$continuous1_data))
rownames(mat_col) <- colnames(dat1)
dat1 <- dat1[, order(as.numeric(simdata$cluster))]
gp_dat1 <- pheatmap(dat1,
  main = "View 1",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
  treeheight_row = 0,
  treeheight_col = 0,
  scale = "row",
  show_colnames = FALSE,
  show_rownames = FALSE,
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  annotation_col = mat_col,
  annotation_colors = mat_colors
)

dat2 <- data.frame(t(simdata$continuous2_data))
rownames(mat_col) <- colnames(dat2)
dat2 <- dat2[, order(as.numeric(simdata$cluster))]
gp_dat2 <- pheatmap(dat2,
  main = "View 2",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
  treeheight_row = 0,
  treeheight_col = 0,
  scale = "row",
  show_colnames = FALSE,
  show_rownames = FALSE,
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  annotation_col = mat_col,
  annotation_colors = mat_colors
)

dat3 <- data.frame(t(simdata$count_data))
rownames(mat_col) <- colnames(dat3)
dat3 <- dat3[, order(as.numeric(simdata$cluster))]
gp_dat3 <- pheatmap(dat3,
  main = "View 3",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
  treeheight_row = 0,
  treeheight_col = 0,
  scale = "row",
  show_colnames = FALSE,
  show_rownames = FALSE,
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  annotation_col = mat_col,
  annotation_colors = mat_colors
)

dat4 <- data.frame(t(simdata$binary_data))
rownames(mat_col) <- colnames(dat4)
dat4 <- dat4[, order(as.numeric(simdata$cluster))]
gp_dat4 <- pheatmap(dat4,
  main = "View 4",
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
  treeheight_row = 0,
  treeheight_col = 0,
  scale = "row",
  show_colnames = FALSE,
  show_rownames = FALSE,
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  annotation_col = mat_col,
  annotation_colors = mat_colors
)
library(ggplotify)
gp_dat1 <- as.ggplot(gp_dat1)
gp_dat2 <- as.ggplot(gp_dat2)
gp_dat3 <- as.ggplot(gp_dat3)
gp_dat4 <- as.ggplot(gp_dat4)
```


```{r fig.width = 16, fig.height=8}
library(gridExtra)
grid.arrange(gp_dat1, gp_dat2, gp_dat3, gp_dat4, nrow = 1)
```



## Using iClusterVB


### Pre-processing data

```{r}
binary_data_new <- simdata$binary_data
binary_data_new[binary_data_new == 0] <- 2 # re-code the categorical variable
```

### Specifying data and the corresponding distributions to analyze

```{r}
mydata <- list(simdata$continuous1_data, simdata$continuous2_data, simdata$count_data, binary_data_new)
dist <- c("gaussian", "gaussian", "poisson", "multinomial")

# Use "gaussian" for continuous data, "poisson"  for count data, and "multinomial" for binary data
```


### Fitting the model

```{r}
K_max <- 8 # This is the maximum number of clusters
set.seed(41)
fit.iClusterVB <- iClusterVB(
  mydata = mydata,
  dist = dist,
  initial_method = "VarSelLCM", # default initilization method
  VS_method = 1,
  K = K_max,
  max_iter = 200,
  per = 100
)
```

### Making a plot for ELBO (evidence lower bound) by iteration

```{r}
iter <- fit.iClusterVB$iter
elbo <- fit.iClusterVB$elbo[1:iter]
plot(1:iter, elbo,
  type = "o",
  lwd = 2,
  xlab = "Iteration",
  ylab = "ELBO"
)
```

### Printing the cluster proportion

```{r}
# ppi is a vector with a length of K_max
# note that the estimated ppi suggested that
# only clusters 2, 3, 4 and 7 are non-empty,
# each of these clusters yielded an estimated cluster proportion of 0.25.
print(fit.iClusterVB$model_parameter$ppi)
```
### Printing the cluster membership 

```{r}
print(fit.iClusterVB$cluster)
table(fit.iClusterVB$cluster)
```


### Using a different (k-prototype) initializing method for cluster membership 

```{r}
set.seed(41)
fit.iClusterVB.kproto <- iClusterVB(
  mydata = mydata,
  dist = dist,
  initial_method = "kproto",
  VS_method = 1,
  K = K_max,
  max_iter = 200,
  per = 100
)
```



#### Making a plot for ELBO (evidence lower bound) by iteration


```{r}
iter <- fit.iClusterVB.kproto$iter
elbo <- fit.iClusterVB.kproto$elbo[1:iter]
plot(1:iter, elbo,
  type = "o",
  lwd = 2,
  xlab = "Iteration",
  ylab = "ELBO"
)
```

#### Printing the cluster membership 

```{r}
table(fit.iClusterVB.kproto$cluster)
```


### Creating a barplot for the cluster proportion

```{r}
library(ggplot2)
K <- length(table(fit.iClusterVB$cluster)) # number of clusters
N <- fit.iClusterVB$data_summary$N # sample size
df <- data.frame(cbind(cluster = 1:K, prop = round(as.numeric(table(fit.iClusterVB$cluster) / N), 2)))
df$cluster <- factor(df$cluster, label = paste0("Cluster ", 1:K))
gp_prop <- ggplot(df, aes(
  x = cluster, y = prop,
  color = cluster, fill = cluster,
  label = scales::percent(prop)
)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_bw() +
  ggtitle(" ") +
  theme(
    legend.position = "none", legend.title = element_blank(),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    axis.text.x = element_text(angle = 0),
    strip.text.x = element_text(size = 15, angle = 0),
    strip.text.y = element_text(size = 15, face = "bold")
  ) +
  geom_text(
    position = position_dodge(.9),
    vjust = -0.5,
    size = 5
  ) +
  xlab("") +
  ylab("Percentage (%)") +
  scale_y_continuous(labels = scales::percent, limit = c(0, 1.05))
```


```{r}
gp_prop
```




### Extract the posterior probability of feature inclusion

```{r}
fit.iClusterVB$model_parameters$rho[[1]] # view 1
fit.iClusterVB$model_parameters$rho[[2]] # view 2
fit.iClusterVB$model_parameters$rho[[3]] # view 3
fit.iClusterVB$model_parameters$rho[[4]] # view 4
```

### Visualizing the clusters and selected features
```{r}
dat1 <- data.frame(varid = 1:length(fit.iClusterVB$model_parameters$rho[[1]]), rho = t(fit.iClusterVB$model_parameters$rho[[1]]))
dat2 <- data.frame(varid = 1:length(fit.iClusterVB$model_parameters$rho[[2]]), rho = t(fit.iClusterVB$model_parameters$rho[[2]]))
dat3 <- data.frame(varid = 1:length(fit.iClusterVB$model_parameters$rho[[3]]), rho = t(fit.iClusterVB$model_parameters$rho[[3]]))
dat4 <- data.frame(varid = 1:length(fit.iClusterVB$model_parameters$rho[[4]]), rho = t(fit.iClusterVB$model_parameters$rho[[4]]))
```

```{r}
library(ggplot2)
gp_rho1 <- ggplot(dat1, aes(x = reorder(varid, rho), y = rho, fill = rho)) +
  geom_bar(stat = "identity") +
  ggtitle("View 1") +
  ylim(c(0, 1)) +
  geom_hline(yintercept = 0.5, color = "red", linewidth = 1, linetype = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 1),
    axis.title = element_text(size = 15),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1),
    strip.text.x = element_text(size = 1, angle = 0),
    strip.text.y = element_text(size = 15, face = "bold")
  ) +
  xlab(" ") +
  ylab("Probability of Inclusion") +
  scale_fill_gradientn(colors = topo.colors(2), limits = c(0, 1)) +
  coord_polar()
gp_rho2 <- ggplot(dat2, aes(x = reorder(varid, rho), y = rho, fill = rho)) +
  geom_bar(stat = "identity") +
  ggtitle("View 2") +
  ylim(c(0, 1)) +
  geom_hline(yintercept = 0.5, color = "red", linewidth = 1, linetype = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 1),
    axis.title = element_text(size = 15),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1),
    strip.text.x = element_text(size = 1, angle = 0),
    strip.text.y = element_text(size = 15, face = "bold")
  ) +
  xlab(" ") +
  ylab("Probability of Inclusion") +
  scale_fill_gradientn(colors = topo.colors(2), limits = c(0, 1)) +
  coord_polar()
gp_rho3 <- ggplot(dat3, aes(x = reorder(varid, rho), y = rho, fill = rho)) +
  geom_bar(stat = "identity") +
  ggtitle("View 3") +
  ylim(c(0, 1)) +
  geom_hline(yintercept = 0.5, color = "red", linewidth = 1, linetype = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 1),
    axis.title = element_text(size = 15),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1),
    strip.text.x = element_text(size = 1, angle = 0),
    strip.text.y = element_text(size = 15, face = "bold")
  ) +
  xlab(" ") +
  ylab("Probability of Inclusion") +
  scale_fill_gradientn(colors = topo.colors(2), limits = c(0, 1)) +
  coord_polar()
gp_rho4 <- ggplot(dat4, aes(x = reorder(varid, rho), y = rho, fill = rho)) +
  geom_bar(stat = "identity") +
  ggtitle("View 4") +
  ylim(c(0, 1)) +
  geom_hline(yintercept = 0.5, color = "red", linewidth = 1, linetype = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 1),
    axis.title = element_text(size = 15),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1),
    strip.text.x = element_text(size = 1, angle = 0),
    strip.text.y = element_text(size = 15, face = "bold")
  ) +
  xlab(" ") +
  ylab("Probability of Inclusion") +
  scale_fill_gradientn(colors = topo.colors(2), limits = c(0, 1)) +
  coord_polar()
```
```{r fig.width=8, fig.height=16 }
library(cowplot)
plot_grid(gp_rho1,
  gp_rho2,
  gp_rho3,
  gp_rho4,
  labels = c("(A)", "(B)", "(C)", "(D)"), nrow = 2, align = "hv"
)
```




